Formalizing the Kolmogorov extension Theorem in Lean

### 2023-09-11 Manuscript/proof_outline.tex
make small changes, add todos

@@

\begin{abstract}
  We present a formalization of the Kolmogorov extension theorem,
  which is a main building block in construction the (distribution of)
  stochastic processes with arbitrary index sets in Lean. Our
  approach is based on mathlib, the mathematical library for Lean. On
  our way, we provide a formalization of the Caratheodory extension
  theorem, which allows to extend a set function on a semi-ring to a
  proper measure on the $\sigma$-algebra generated by the semi-ring. A
  slight generalization of the classical Kolmogorov extension theorem
  allows us to construct the (distribution of) a stochastic process on
  a complete and separable pseudo-metric space (rather than a metric
  space).
\end{abstract}

@@

\section{Introduction}
One of the main building blocks of modern probability theory are stochastic processes, which are usually defined as any collection of random variables -- $(X_t)_{t\in \iota}$ with $X_t$ taking values in some $\alpha_t$ for all $t\in\iota$ -- defined on some joint probability space $(\Omega, \mathcal A, \mathbf P)$.  (As usual, we will refer to $\iota$ as the index set of times.) In order to study such processes, it is fundamental to talk about their joint distribution, i.e.\ a probability distribution on the product set $\prod_{t\in\iota} \alpha_t$. The usual approach to construct (the distribution of) a stochastic process works as follows: describe properties of the distribution of the stochastic process $P_J$ at some arbitrary but finite number of times $J = \{t_1,...,t_n\} \subseteq \iota$. The resulting family of probability measures $(P_J)_{J \subseteq \iota \text{ finite}}$ has to be {\em projective} in the sense that the projection of $P_J$ to $H\subseteq J$ has to be equal to $P_H$. In other words, when describing the distribution of the stochastic process at all times in $J$, and then forgetting (i.e.\ projecting) all properties for times in $J\setminus H$, results in the description of properties at times in $H$. One may then ask if this already gives a complete description of the process for all times. For uncountable $\iota$, e.g. $\iota = [0,\infty)$, one is tempted to be pessimistic at first sight since measures -- which describe the distribution of the stochastic process -- usually only deal well with a countable number of measurable events. However, it is the achievement of Kolmogorov that the finite-dimensional distributions in fact provide a unique description of the distribution of a stochastic process, as long as the underlying family of state spaces $(\alpha_t)_{t\in\iota}$ is nice enough (Polish, i.e.\ a separable topological space which can be metrized by a complete metric, for example). Assuming a family of topological space $(\alpha_t)_{t\in\iota}$, this distribution is a measure on the product-$\sigma$-field $\mathcal F :=\bigotimes_{t\in\iota} \mathcal B(\alpha_t)$ (where $\mathcal B(\alpha_t)$ is the Borel $\sigma$-algebra on $\alpha_t$). Here, $\mathcal F$ is generated by finite projections and hence any element of $\mathcal F$ may only depend on at most countably many $t\in \iota$, making this a rather coarse $\sigma$-algebra. (In particular, note that this is not the Borel $\sigma$-algebra of the product topology for infinite $\iota$.) The result responsible for this insight is usually denoted the {\em Kolmogorov extension theorem}, formulated in \cite{kolmogoroff1933grundbegriffe}. We note that a version of the extension result was proved by Daniell in the 1930s, but this paper was not acknowledged by the probabilists of that time \cite{aldrich2007but}. Due to his contribution, the theorem is often called the {\em Daniell-Kolmogorov extension}.

The goal of our contribution is a formalization of the proof of that theorem in {\tt Lean}. Here, we are going to use \leanline{Lean4}, since \leanline{mathlib} \cite{mathlib2020lean}, the mathematical library of \leanline{Lean} has recently been ported to \leanline{Lean4}. Subsequently, we aim to make a pull request for the publication of our code in \leanline{mathlib}. The Kolmogorov extension theorem is on the interface between measure theory and probability theory. Here, we rely on a decent amount of formalized mathematics in the measure-theory part of {\tt mathlib} (outer measures, above all), while not requiring any specific previous formalization of probability theory. (In fact, most of our results are formulated in terms of finite rather than probability measures.)

At first sight, it might be surprising that martingales, a certain class of stochastic process, have already been formalized in Lean \cite{ying2023formalization}, although the Kolmogorov extension theorem (on the existence of stochastic processes) is only available by our contribution. Note, however, that a martingale $(X_t)_{t\in \iota}$ is defined as a family of random variables (satisfying some properties) on a fixed probability space $(\Omega, \mathcal F, \mathbf P)$, while the Kolmogorov extension theorem is on the construction of a probability (or finite) measure $\mathbf P$, on which we can define random variables (with certain properties). So, our work complements \cite{ying2023formalization} in the sense that probability spaces $(\Omega, \mathcal F, \mathbf P)$ on which you can define a martingale exist. However, we do not give any specific example for the construction of a specific instance of such a space.

The main steps in our construction, which were previously missing in {\tt mathlib}, are (more mathematical details are below, and in standard textbooks on Probability Theory, e.g.\ \cite{Klenke2013, Kallenberg2020}):
\begin{enumerate}
\item a formalization of (the set-system of) semi-rings (see Definition~\ref{def:semi} below); 
\item adding some main lemmas to measure theory, e.g.\ continuity of a measure at the emptyset (see e.g.\ Lemma~\ref{l:halbextRing}, Proposition~\ref{P:stetigmass}, Lemma~\ref{l:stetigcompact} below);
\item a proof that a single probability measure on a Polish space is inner regular with respect to compact sets (see Definition~\ref{def:innerreg}; in fact, we proved a slight generalization using pseudo-metric spaces, found in Lemma~\ref{L:relcoPol} below); 
\item the classical Carathéodory extension theorem, providing us a candidate for the measure which we want to construct (see Theorem~\ref{T:masseind}); 
\item the Kolmogorov extension theorem, as based on the previous steps (See Theorem~\ref{T1}).
\end{enumerate}
The Kolmogorov extension theorem has been previously formalized in Isabelle/HOL \cite{Immler2012}. This formalization only works on Polish spaces (rather than on spaces where every finite measure is inner regular with respect to compact sets, see below), and only in the case where all $\alpha_i$'s are identical. However, type theory as the basis of the formalization, as well as several mathematical concepts (e.g.\ projective families of measures or the continuity of a measure in the emptyset) are the same in both formalisations in these two interactive theorem provers.

@@

\noindent
{\em Instances of stochastic processes:} An obvious application of Kolmogorov's extension theorem is the construction of basic stochastic processes like Markov chains \cite[Section~11]{Kallenberg2020}, the Poisson process \cite[Section~13]{Kallenberg2020} and Brownian motion \cite[Section~14]{Kallenberg2020}. (Note that other Gaussian processes -- indexed by $[0, \infty)$ or $\mathbb R$ -- might as well be constructed the same way. In addition, fields like the Gaussien free field indexed by $\mathbb R^d$ \cite{sheffield2007gaussian} can also be given.) For these tasks, we would have to define the finite-dimensional distributions (Poisson and normally distributed, respectively), and apply the extension theorem. This task requires the formalization of multi-dimensional Poisson and normal distributions, which -- in textbooks -- is usually done using characteristic functions. Since these are not yet part of {\tt mathlib}, we postpone this task to the future.

\noindent
{\em Sample-path properties:} The Kolmogorov extension gives the existence of a distribution of a stochastic process $(X_t)_{t\in\iota}$ with certain properties. Extra work is needed in order to show that -- on the same probability space -- we can as well define a version (i.e.\ another process $(Y_t)_{t\in\iota}$ with $\mathbf P(X_t = Y_t) = 1$ for all $t\in\iota$) which is right-continuous with left limits (for the Poisson process) or continuous (for Brownian motion), respectively. For the former, this follows from some general principles of Markov processes (e.g.\ \cite[Theorem 4.3.6]{EthierKurtz1986}). For the latter, this requires formalization of the Kolmogorov-Chentsov criterion 
\cite[Theorem~4.23]{Kallenberg2020}.

\noindent
{\em Theorem of Ionescu-Tulcea and related results:} While Kolmogorov's extension Theorem gives a result for arbitrary $\iota$, but requires some properties of the family $(\alpha_t)_{t\in\iota}$, the Theorem of Ionescu-Tulcea \cite[Theorem 8.24]{Kallenberg2020} can only deal with countable $\iota$, but has no restrictions on $(\alpha_t)_{t\in\iota}$. The proof of the latter, however, uses transition kernels (and an inductive construction of a content, which can be extended to a measure) rather than projective families. However, continuity of a measure at $\emptyset$ is used as well and a formalization would complement our work. In addition, using the Theorem of Ionescu-Tulcea one can define infinitely many independent random variables (arbitrary $\iota$) on a joint probability space \cite[Theorem 8.24]{Kallenberg2020} without requirements on $(\alpha_t)_{t\in\iota}$. All of these results can be used in order to construct countably many independent random variables, which are frequently used in concrete constructions in probability theory. Examples are (one direction of) the Borel-Cantelli-Lemma \cite[Theorem 4.28]{Kallenberg2020}, Kolmogorov's 0-1-law \cite[Theorem 4.13]{Kallenberg2020}, random walks \cite[Chapter 12]{Kallenberg2020}, branching processes  \cite[Chapter 13]{Kallenberg2020}, and percolation  \cite[Chapter 2.4]{Klenke2013}, to name just a few. While the first two are already formalized in \leanline{mathlib}, it is only clear due to the Kolmorov extension theorem (or the Theorem of Ionesu-Tulcea) that infinitely many independent random variables can indeed be defined on a joint probability space. In this sense, our contribution is also important for giving more sense to already formalised parts of \leanline{mathlib}.

@@

\begin{definition}[Metric and topological spaces]
  \begin{enumerate}
  \item \sloppy A pseudo-metric on $\alpha$ is a symmetric map $r : \alpha
    \times \alpha \to [0,\infty)$ satisfying the triangle inequality,
      i.e.\ $r(x,z) \leq r(x,y) + r(y,z)$ for all $x,y,z\in\alpha$.
      \\ If $r$ also satisfies $r(x,y) = 0$ iff $x=y$, we call it a
      metric.  \\ If $r : \alpha \times \alpha \to [0,\infty]$
      (i.e.\ $r(x,y) = \infty$ is allowed), we call $r$ an extended
      (pseudo-)metric.
    \item Let $r$ be an (extended pseudo-)metric. A sequence $x_1,
      x_2,... \in \alpha$ is called Cauchy if for all $\varepsilon>0$
      there is $N\in\mathbb N$ such that $r(x_n, x_m) < \varepsilon$
      for all $m,n > N$. The (extended pseudo-) metric is called
      complete if every Cauchy-sequency has a limit in
      $\alpha$.
    \item Some\footnote{We denote by $2^\alpha$ the power set of
      $\alpha$, i.e.\ the set of all subsets of $\alpha$.} $\mathcal O
      \subseteq 2^\alpha$ is called a topology, if it satisfies (i)
      $\emptyset, \alpha \in \mathcal O$, (ii) $\mathcal O$ is a
      $\pi$-system, i.e.\ it is stable under finite intersections,
      i.e.\ if $A, B \in \mathcal O$, then $A \cap B \in \mathcal O$,
      and (iii) $\mathcal O$ is stable under arbitrary unions,
      i.e.\ if $A_i \in \mathcal O$ for all $i\in\iota$ and $\iota$ is
      arbitrary, then $\bigcup_{i\in \iota} A_i \in \mathcal O$.
  \end{enumerate}
\end{definition}

\begin{remark}[Metric and topological spaces in {\tt mathlib}] \mbox{}
  \begin{enumerate}
    \item \sloppy In \leanline{Lean}, we have the \leanline{class
      PseudoEMetricSpace (α : Type u) extends EDist}, where
      \leanline{EDist} comes with an extended distance \leanline{edist
        : α → α → ENNReal}, and the extension to
      \leanline{PseudoEMetricSpace} consists of the components,
      \leanline{edist_self}, \leanline{edist_comm},
      \leanline{edist_triangle} providing the properties of the
      pseudo-metric, and \leanline{toUniformSpace},
      \leanline{uniformity_edist} defining a uniform space from the
      extended pseudo-metric. Such a space does not come with a
      metric, but with a filter on $\alpha \times \alpha$, which
      describes which points in $\alpha$ are near. For example, the
      diagonal of $\alpha \times \alpha$ is a subset of all sets in
      the uniformity; see \cite{james2013topologies} for details. We
      note that a uniform space with a countably generated uniformity
      filter is pseudometrizable, i.e.\ there exists a
      pseudo-metric-space structure that generates the same
      uniformity; see \leanline{UniformSpace.pseudoMetrizableSpace},
      which formalizes a result stated in
      \cite{melikhov2011metrizable}.
    \item \sloppy For topological spaces, we have \leanline{class
      TopologicalSpace(α : Type u) : Type u}, which comes with
      \leanline{IsOpen : Prop}, \leanline{isOpen_univ : IsOpen
        Set.univ}, \leanline{isOpen_inter} and
      \leanline{isOpen_sUnion}, which are exactly the properties of a
      topological space described above.
    \item Any (extended pseudo-)metric on $\alpha$ defines a topology,
      namely the topology generated by $\mathcal H := \{\{y:
      r(x,y)<\varepsilon\}: x \in \alpha, \varepsilon \in
      (0,\infty)\}$. Actually, in \leanline{mathlib}, as we have seen
      above, every extended pseudo-metric defines a uniform space, and
      a uniform space is a \leanline{class UniformSpace(α : Type u)
        extends TopologicalSpace , UniformSpace.Core} (where the
      uniformity is defined in \leanline{UniformSpace.Core}), which
      connects the \leanline{uniformity} with the topology using
      \leanline{∀ (s : Set α), IsOpen s ↔ ∀ (x : α), x ∈ s → {p |
          p.fst = x → p.snd ∈ s} ∈ uniformity}. In particular, a
      uniform space defines a topological space, which can be used by
      the typeclass system.
  \end{enumerate}
\end{remark}

@@

\begin{remark}[Measur(abl)e spaces in \leanline{mathlib}]
  \begin{enumerate}
  \item \sloppy The \leanline{class MeasurableSpace α} is very similar
    to a topological space in \leanline{mathlib}, since it comes with
    \leanline{MeasurableSet' : Set α → Prop},
    \leanline{measurableSet_empty : MeasurableSet' ∅},
    \leanline{measurableSet_compl : (s : Set α) → MeasurableSet' s →
      MeasurableSet' sᶜ}, and \leanline{measurableSet_iUnion},
    i.e.\ properties (i)-(iii) from Definition~\ref{def:sigma}.1.
  \item We are using outer measures in our construction. In
    \leanline{mathlib}, this is \leanline{structure
      MeasureTheory.OuterMeasure α} which is a function
    \leanline{measureOf : Set α → ℝ≥0∞} such that \leanline{measureOf
      ∅ = 0} (i.e.\ the empty set has measure~$0$), \leanline{∀ {s₁ s₂
        : Set α}, s₁ ⊆ s₂ → measureOf s₁ ≤ measureOf s₂}
    (i.e.\ monotonicity) and \leanline{∀ (s : ℕ → Set α), measureOf (⋃
      (i : ℕ), s i) ≤ ∑' (i : ℕ), measureOf (s i)} (which we call
    $\sigma$-subadditivity). \\ Note that an outer measure is defined
    on $2^\alpha$, whereas a measure is only defined on a subset (the
    $\sigma$-algebra of measurable sets). More precisely: A
    \leanline{MeasureSpace α} extends a \leanline{MeasurableSpace α}
    by a \leanline{Measure α} (which is usually called
    \leanline{volume}). For the measure, we have: \leanline{structure
      MeasureTheory.Measure α [MeasurableSpace α] extends
      MeasureTheory.OuterMeasure}. Here $\sigma$-additivity of a
    \leanline{Measure α} reads
\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  m_iUnion : ∀ ⦃f : ℕ → Set α⦄,
    (∀ (i : ℕ), MeasurableSet (f i)) →
    Pairwise (Disjoint on f) →
    ↑toOuterMeasure (⋃ (i : ℕ), f i) =
    ∑' (i : ℕ), ↑toOuterMeasure (f i)
\end{minted}    
  \end{enumerate}
\end{remark}

@@

\noindent
With this, we can define the projetive family as follows. Note that
\leanline{Lean} automatically uses the subset relation as
\leanline{[Preorder (Finset ι)]} when \leanline{isProjective} is
called.

@@

\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  noncomputable def
    projectiveLimitWithWeakestHypotheses
    [∀ i, PseudoEMetricSpace (α i)]
    [∀ i, BorelSpace (α i)]
    [∀ i, TopologicalSpace.SecondCountableTopology (α i)]
    [∀ i, CompleteSpace (α i)] [Nonempty (∀ i, α i)]
    (P : ∀ J : Finset ι, Measure (∀ j : J, α j))
    [∀ i, IsFiniteMeasure (P i)]
    (hP : IsProjectiveMeasureFamily P) :
    Measure (∀ i, α i)
\end{minted}

We note that we extend the standard assumption that all $\alpha_t$ are
separable, complete metric spaces (or Polish, i.e.\ separable and
metrizable through a complete metric) to cover the case of extended
pseude-metric spaces. Such spaces do not satisfy the frequently used
Hausdorff (or t2) property, i.e.\ there can be $x\neq y$ such that all
open balls around $x$ and $y$ overlap.

@@

\noindent
Using \leanline{variable {α : Type _} {C : Set (Set α)} {s t : Set α}
}, we define a semi-ring as follows:

@@

\noindent
The existence-statement of the above lemma actually gives rise to a
definition. Here, we use \leanline{Exists.choose} in order to extract
an element from an $\exists$-statement. In addition, we do not allow
$\emptyset \in \mathcal K$ without loss of generality:

@@

>>>>

\usepackage[textsize=tiny]{todonotes}

@@

\begin{abstract}
  We present a formalization in Lean of the Kolmogorov extension theorem,
  which is a main building block in the construction of the (distribution of)
  stochastic processes with arbitrary index sets. Our
  approach is based on mathlib, the mathematical library for Lean. On
  our way, we provide a formalization of the Caratheodory extension
  theorem, which allows to extend a set function on a semi-ring to a
  proper measure on the $\sigma$-algebra generated by the semi-ring. A
  slight generalization of the classical Kolmogorov extension theorem
  allows us to construct the (distribution of) a stochastic process on
  a complete and separable pseudo-metric space rather than a metric
  space.
\end{abstract}

@@

\todo[inline]{Clearly differentiate refs to mathlib vs refs to new lemmas?}

\todo[inline]{Mention a precise commit of Mathlib}

\todo[inline]{write about sets or types?}

\todo[inline]{\leanline{∀ j : J, α j} or \leanline{(j : J) -> α j} ?}

\todo[inline]{omit all noncomputable? Otherwise we need to explain what it means}

\section{Introduction}
One of the main building blocks of modern probability theory are stochastic processes, which are usually defined as any collection of random variables -- $(X_t)_{t\in \iota}$ with $X_t$ taking values in some $\alpha_t$ for all $t\in\iota$ -- defined on some joint probability space $(\Omega, \mathcal A, \mathbf P)$.  (As usual, we will refer to $\iota$ as the index set of times.) In order to study such processes, it is fundamental to talk about their joint distribution, i.e.\ a probability distribution on the product set $\prod_{t\in\iota} \alpha_t$. The usual approach to construct (the distribution of) a stochastic process works as follows: describe properties of the distribution of the stochastic process $P_J$ at some arbitrary but finite number of times $J = \{t_1,...,t_n\} \subseteq \iota$.
The resulting family of probability measures $(P_J)_{J \subseteq \iota \text{ finite}}$ has to be {\em projective} in the sense that the projection of $P_J$ to $H\subseteq J$ has to be equal to $P_H$. In other words, when describing the distribution of the stochastic process at all times in $J$, and then forgetting all properties for times in $J\setminus H$, results in the description of properties at times in $H$. One may then ask if this already gives a complete description of the process for all times.
For uncountable $\iota$, e.g. $\iota = [0,\infty)$, one is tempted to be pessimistic at first sight since measures -- which describe the distribution of the stochastic process -- usually only deal well with a countable number of measurable events. However, it is the achievement of Kolmogorov that the finite-dimensional distributions in fact provide a unique description of the distribution of a stochastic process, as long as the underlying family of state spaces $(\alpha_t)_{t\in\iota}$ is nice enough (Polish, i.e.\ a separable topological space which can be metrized by a complete metric, for example).
Assuming a family of topological spaces $(\alpha_t)_{t\in\iota}$, this distribution is a measure on the product-$\sigma$-field $\mathcal F :=\bigotimes_{t\in\iota} \mathcal B(\alpha_t)$ (where $\mathcal B(\alpha_t)$ is the Borel $\sigma$-algebra on $\alpha_t$). Here, $\mathcal F$ is generated by finite projections and hence any element of $\mathcal F$ may only depend on at most countably many $t\in \iota$, making this a rather coarse $\sigma$-algebra. (In particular, note that this is not the Borel $\sigma$-algebra of the product topology for infinite $\iota$.) The result responsible for this insight is usually denoted the {\em Kolmogorov extension theorem}, formulated in \cite{kolmogoroff1933grundbegriffe}. We note that a version of the extension result was proved by Daniell in the 1930s, but this paper was not acknowledged by the probabilists of that time \cite{aldrich2007but}. Due to his contribution, the theorem is often called the {\em Daniell-Kolmogorov extension}.

The goal of our contribution is a formalization of the proof of that theorem in {\tt Lean}. Here, we are going to use \leanline{Lean4}, since \leanline{mathlib} \cite{mathlib2020lean}, the mathematical library of \leanline{Lean} has recently been ported to \leanline{Lean4}.
%Subsequently, we aim to make a pull request for the publication of our code in \leanline{mathlib}.
The Kolmogorov extension theorem is on the interface between measure theory and probability theory. Here, we rely on a decent amount of formalized mathematics in the measure-theory part of {\tt mathlib} (outer measures, above all), while not requiring any specific previous formalization of probability theory. (In fact, most of our results are formulated in terms of finite rather than probability measures.)

At first sight, it might be surprising that martingales, a certain class of stochastic process, have already been formalized in Lean \cite{ying2023formalization}, although the Kolmogorov extension theorem (on the existence of stochastic processes) is only available by our contribution. Note, however, that a martingale $(X_t)_{t\in \iota}$ is defined as a family of random variables (satisfying some properties) on a fixed probability space $(\Omega, \mathcal F, \mathbf P)$, while the Kolmogorov extension theorem is on the construction of a probability (or finite) measure $\mathbf P$, on which we can define random variables (with certain properties). So, our work complements \cite{ying2023formalization} in the sense that probability spaces $(\Omega, \mathcal F, \mathbf P)$ on which you can define a martingale exist. However, we do not give any specific example for the construction of a specific instance of such a space\todo{could we give one?}.

The main steps in our construction, which were previously missing in {\tt mathlib}, are (more mathematical details are below, and in standard textbooks on Probability Theory, e.g.\ \cite{Klenke2013, Kallenberg2020}):
\begin{enumerate}
\item a formalization of (the set-system of) semi-rings (see Definition~\ref{def:semi} below); 
\item adding some main lemmas to measure theory, e.g.\ continuity of a measure at the emptyset\todo{for a measure, this is already in mathlib?} (see e.g.\ Lemma~\ref{l:halbextRing}, Proposition~\ref{P:stetigmass}, Lemma~\ref{l:stetigcompact} below);
\item a proof that a single probability measure on a Polish space is inner regular with respect to compact sets (see Definition~\ref{def:innerreg}; in fact, we proved a slight generalization using pseudo-metric spaces, found in Lemma~\ref{L:relcoPol} below); 
\item the classical Carathéodory extension theorem, providing us with a candidate for the measure which we want to construct (see Theorem~\ref{T:masseind}); 
\item the Kolmogorov extension theorem, as based on the previous steps (See Theorem~\ref{T1}).
\end{enumerate}
The Kolmogorov extension theorem has been previously formalized in Isabelle/HOL \cite{Immler2012}. This formalization only works on Polish spaces (rather than on spaces where every finite measure is inner regular with respect to compact sets, see below), and only in the case where all $\alpha_i$'s are identical\todo{Lean has dependent types, Isabelle doesn't}. However, type theory as the basis of the formalization, as well as several mathematical concepts (e.g.\ projective families of measures or the continuity of a measure in the emptyset) are the same in both formalisations in these two interactive theorem provers.

@@

\noindent
{\em Instances of stochastic processes:} An obvious application of Kolmogorov's extension theorem is the construction of basic stochastic processes like Markov chains \cite[Section~11]{Kallenberg2020}, the Poisson process \cite[Section~13]{Kallenberg2020} and Brownian motion \cite[Section~14]{Kallenberg2020}.
Other Gaussian processes -- indexed by $[0, \infty)$ or $\mathbb R$ -- might as well be constructed the same way. In addition, fields like the Gaussien free field indexed by $\mathbb R^d$ \cite{sheffield2007gaussian} can also be given.
For these tasks, we would have to define the finite-dimensional distributions (Poisson and normally distributed, respectively), and apply the extension theorem. This task requires the formalization of multi-dimensional Poisson and normal distributions, which -- in textbooks -- is usually done using characteristic functions. Since these are not yet part of {\tt mathlib}, we postpone this task to the future.

\noindent
{\em Sample-path properties:} The Kolmogorov extension gives the existence of a distribution of a stochastic process $(X_t)_{t\in\iota}$ with certain properties. Extra work is needed in order to show that -- on the same probability space -- we can as well define a version (i.e.\ another process $(Y_t)_{t\in\iota}$ with $\mathbf P(X_t = Y_t) = 1$ for all $t\in\iota$) which is right-continuous with left limits (for the Poisson process) or continuous (for Brownian motion).
For the former, this follows from some general principles of Markov processes (e.g.\ \cite[Theorem 4.3.6]{EthierKurtz1986}). For the latter, this requires formalization of the Kolmogorov-Chentsov criterion 
\cite[Theorem~4.23]{Kallenberg2020}.

\noindent
{\em Theorem of Ionescu-Tulcea and related results:} While Kolmogorov's extension Theorem gives a result for arbitrary $\iota$, but requires some properties of the family $(\alpha_t)_{t\in\iota}$, the Theorem of Ionescu-Tulcea \cite[Theorem 8.24]{Kallenberg2020} can only deal with countable $\iota$, but has no restrictions on $(\alpha_t)_{t\in\iota}$. The proof of the latter, however, uses transition kernels (and an inductive construction of a content, which can be extended to a measure) rather than projective families. However, continuity of a measure at $\emptyset$\todo{already in mathlib} is used as well and a formalization would complement our work. In addition, using the Theorem of Ionescu-Tulcea one can define infinitely many independent random variables (arbitrary $\iota$) on a joint probability space \cite[Theorem 8.24]{Kallenberg2020} without requirements on $(\alpha_t)_{t\in\iota}$. All of these results can be used in order to construct countably many independent random variables, which are frequently used in concrete constructions in probability theory. Examples are (one direction of) the Borel-Cantelli-Lemma \cite[Theorem 4.28]{Kallenberg2020}, Kolmogorov's 0-1-law \cite[Theorem 4.13]{Kallenberg2020}, random walks \cite[Chapter 12]{Kallenberg2020}, branching processes  \cite[Chapter 13]{Kallenberg2020}, and percolation  \cite[Chapter 2.4]{Klenke2013}, to name just a few. While the first two are already formalized in \leanline{mathlib}, it is only clear due to the Kolmorov extension theorem (or the Theorem of Ionesu-Tulcea) that infinitely many independent random variables can indeed be defined on a joint probability space. In this sense, our contribution is also important for giving more sense to already formalised parts of \leanline{mathlib}.

@@

\begin{definition}[Metric and topological spaces]
  \begin{enumerate}
  \item \sloppy A pseudo-metric on $\alpha$ is a symmetric map $r : \alpha
    \times \alpha \to [0,\infty)$ satisfying the triangle inequality,
      i.e.\ $r(x,z) \leq r(x,y) + r(y,z)$ for all $x,y,z\in\alpha$.
      \\ If $r$ also satisfies $r(x,y) = 0$ iff $x=y$, we call it a
      metric.  \\ If $r : \alpha \times \alpha \to [0,\infty]$
      (i.e.\ $r(x,y) = \infty$ is allowed), we call $r$ an extended
      (pseudo-)metric.
    \item Let $r$ be an (extended pseudo-)metric. A sequence $x_1,
      x_2,... \in \alpha$ is called Cauchy if for all $\varepsilon>0$
      there is $N\in\mathbb N$ such that $r(x_n, x_m) < \varepsilon$
      for all $m,n > N$. The (extended pseudo-) metric is called
      complete if every Cauchy-sequency has a limit in
      $\alpha$.
    \item Some\footnote{We denote by $2^\alpha$ the power set of
      $\alpha$, i.e.\ the set of all subsets of $\alpha$.} $\mathcal O
      \subseteq 2^\alpha$ is called a topology, if it satisfies (i)
      $\emptyset, \alpha \in \mathcal O$, (ii) $\mathcal O$ is a
      $\pi$-system, i.e.\ it is stable under finite intersections,
      i.e.\ if $A, B \in \mathcal O$, then $A \cap B \in \mathcal O$,
      and (iii) $\mathcal O$ is stable under arbitrary unions,
      i.e.\ if $A_i \in \mathcal O$ for all $i\in\iota$ and $\iota$ is
      arbitrary, then $\bigcup_{i\in \iota} A_i \in \mathcal O$.
  \end{enumerate}
\end{definition}
\todo{reorder that next remark? I'll make a proposal}
\begin{remark}[Metric and topological spaces in {\tt mathlib}] \mbox{}
  \begin{enumerate}
    \item \sloppy In \leanline{Lean}, we have the \leanline{class
      PseudoEMetricSpace (α : Type u) extends EDist}, where
      \leanline{EDist} comes with an extended distance \leanline{edist: α → α → ℝ≥0∞}
      (the notation \leanline{ℝ≥0∞} stands for $[0, \infty]$), and the extension to
      \leanline{PseudoEMetricSpace} consists of the components,
      \leanline{edist_self}, \leanline{edist_comm},
      \leanline{edist_triangle} providing the properties of the
      pseudo-metric, and \leanline{toUniformSpace},
      \leanline{uniformity_edist} defining a uniform space from the
      extended pseudo-metric. Such a space does not come with a
      metric, but with a filter on $\alpha \times \alpha$, which
      describes which points in $\alpha$ are near. For example, the
      diagonal of $\alpha \times \alpha$ is a subset of all sets in
      the uniformity; see \cite{james2013topologies} for details. We
      note that a uniform space with a countably generated uniformity
      filter is pseudometrizable, i.e.\ there exists a
      pseudo-metric-space structure that generates the same
      uniformity; see \leanline{UniformSpace.pseudoMetrizableSpace},
      which formalizes a result stated in
      \cite{melikhov2011metrizable}.
    \item \sloppy For topological spaces, we have \leanline{class TopologicalSpace(α : Type u) : Type u}, which comes with
      \leanline{IsOpen : Prop}, \leanline{isOpen_univ : IsOpen Set.univ}, \leanline{isOpen_inter} and
      \leanline{isOpen_sUnion}, which are exactly the properties of a
      topological space described above.
    \item Any (extended pseudo-)metric on $\alpha$ defines a topology,
      namely the topology generated by $\mathcal H := \{\{y:
      r(x,y)<\varepsilon\}: x \in \alpha, \varepsilon \in
      (0,\infty)\}$. Actually, in \leanline{mathlib}, as we have seen
      above, every extended pseudo-metric defines a uniform space, and
      a uniform space is a \leanline{class UniformSpace(α : Type u)
        extends TopologicalSpace , UniformSpace.Core} (where the
      uniformity is defined in \leanline{UniformSpace.Core}), which
      connects the \leanline{uniformity} with the topology using
      \leanline{∀ (s : Set α), IsOpen s ↔ ∀ (x : α), x ∈ s → {p |
          p.fst = x → p.snd ∈ s} ∈ uniformity}. In particular, a
      uniform space defines a topological space, which can be used by
      the typeclass system.
  \end{enumerate}
\end{remark}

@@

\begin{remark}[Measur(abl)e spaces in \leanline{mathlib}]
  \begin{enumerate}
  \item \sloppy The \leanline{class MeasurableSpace α} is very similar
    to a topological space in \leanline{mathlib}, since it comes with
    \leanline{MeasurableSet' : Set α → Prop},
    \leanline{measurableSet_empty : MeasurableSet' ∅},
    \leanline{measurableSet_compl : (s : Set α) → MeasurableSet' s →
      MeasurableSet' sᶜ}, and \leanline{measurableSet_iUnion},
    i.e.\ properties (i)-(iii) from Definition~\ref{def:sigma}.1.
  \item We are using outer measures in our construction. In
    \leanline{mathlib}, this is \leanline{structure
      MeasureTheory.OuterMeasure α} which is a function
    \leanline{measureOf : Set α → ℝ≥0∞} such that \leanline{measureOf
      ∅ = 0} (i.e.\ the empty set has measure~$0$), \leanline{∀ {s₁ s₂
        : Set α}, s₁ ⊆ s₂ → measureOf s₁ ≤ measureOf s₂}
    (i.e.\ monotonicity) and \leanline{∀ (s : ℕ → Set α), measureOf (⋃
      (i : ℕ), s i) ≤ ∑' (i : ℕ), measureOf (s i)} (which we call
    $\sigma$-subadditivity). \\ Note that an outer measure is defined
    on $2^\alpha$, whereas a measure is only defined on a subset (the
    $\sigma$-algebra of measurable sets). More precisely\todo{I don't understand why all these details are here. We never use MeasureSpace}: A
    \leanline{MeasureSpace α} extends a \leanline{MeasurableSpace α}
    by a \leanline{Measure α} (which is usually called
    \leanline{volume}). For the measure, we have: \leanline{structure
      MeasureTheory.Measure α [MeasurableSpace α] extends
      MeasureTheory.OuterMeasure}. Here $\sigma$-additivity of a
    \leanline{Measure α} reads
\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  m_iUnion : ∀ ⦃f : ℕ → Set α⦄,
    (∀ (i : ℕ), MeasurableSet (f i)) →
    Pairwise (Disjoint on f) →
    ↑toOuterMeasure (⋃ (i : ℕ), f i) =
    ∑' (i : ℕ), ↑toOuterMeasure (f i)
\end{minted}    
  \end{enumerate}
\end{remark}

@@

\noindent
With this, we can define the projetive family as follows. The typeclass inference system of
\leanline{Lean} automatically uses the subset relation to generate
\leanline{[Preorder (Finset ι)]} when \leanline{isProjective} is
called.

@@

\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  noncomputable def
    projectiveLimitWithWeakestHypotheses
    [∀ i, PseudoEMetricSpace (α i)]
    [∀ i, BorelSpace (α i)]
    [∀ i, SecondCountableTopology (α i)]
    [∀ i, CompleteSpace (α i)] [∀ i, Nonempty (α i)]
    (P : ∀ J : Finset ι, Measure (∀ j : J, α j))
    [∀ i, IsFiniteMeasure (P i)]
    (hP : IsProjectiveMeasureFamily P) :
    Measure (∀ i, α i)
\end{minted}

We note that we extend the standard assumption that all $\alpha_t$ are
separable, complete metric spaces (or Polish, i.e.\ separable and
metrizable through a complete metric) to cover the case of extended
pseude-metric spaces. Such spaces do not satisfy the frequently used
Hausdorff (or t2) property, i.e.\ there can be $x\neq y$ such that all
open balls around $x$ and $y$ overlap\todo{it would be nice to explain how the formalization led us to remove that hypothesis}.

@@

\noindent
Any ring is a semi-ring since $A\cap B = A \setminus (A \setminus B)$,
i.e.\ every ring is a $\pi$-system.
Using \leanline{variable {α : Type _} {s t : Set α}
}, we define a semi-ring as follows:

@@

\noindent
The existence-statement of the above lemma actually gives rise to a
definition. Here, we use \leanline{Exists.choose} in order to extract
an element from an $\exists$-statement. In addition, we do not allow
$\emptyset \in \mathcal K$ without loss of generality\todo{that exclusion was useful for something. Find out what that is}:

@@

\todo[inline]{Is this section going into too much detail?}
